Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-03-30 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Improve section about Anderson's variant

\section{A natural deduction proof based on Anderson's axiomatization}
\label{sec:anderson}

@@

%Anderson modification of the axioms in A1 equivalence removed and definition of essence altered. 
%Anderson's implication instead of equivalence. 
\begin{axiom}
\label{A1}
Either a property or its negation is positive, but not both:
$$
\all \varphi. [P(\neg \varphi) \imp \neg P(\varphi)]
$$
\end{axiom}

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}

The proof is otherwise identical to the proof of theorem \ref{T1} in section \ref{sec:newproof} but the equivalence of axiom \ref{A1} does not need to be reduced to an implication by implication elimination.

\begin{definition}[Anderson's alternative Godlike]
\label{D1}
%A \emph{God-like} being possesses all positive properties:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \biimp \nec \varphi(x)]
$$
\end{definition}

\begin{corollary}
\label{C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}

\begin{definition}[Anderson's alternative essence with Scott's additional conjunct]
\label{D2}
%sAn \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties:
$$
\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

%At this point, G\"odel and Scott proceed to prove Theorem 2, which is then used to prove Lemma 1. However, a close inspection of Scott's proof of Lemma 1 reveals that Theorem 2 is an unnecessary detour. Therefore, we do not include it in the new proof.

\begin{definition}[Anderson's alternative where G\"odel's definition of essence is replaced by def. \ref{D2}]
\label{D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
$$
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex y.\varphi(y)]
$$
\end{definition}

\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_6] \quad G(\gamma)$}
\end{prooftree}

@@

\begin{small}
\begin{prooftree}
\AXC{$\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\nec E(\gamma)$}\RightLabel{$\nec_E$ (reflexive frame)}
\UIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

\subsection{Why is the proof of modal collapse prevented?}

The alteration of the definition of essence in Anderson's proof prevents the standard proof of modal collapse \ref{Collapse} 
$$\vdash A \imp \nec A
$$ 
by the additional necessitation, which converts the proof of modal collapse into a proof of a tautology.
$$
\vdash \nec A \imp \nec A
$$

%For all propositions $A$, the following \emph{modal collapse} proposition is not provable with Anderson's alternative definitions with the same proof as G\"odel or Scott's:

@@

The exchange of the implication in the G\"odel-Scott definition of essence for an equivalence and the additional  modal necessitation of the property $A$ prevents the implication of a necessary fact from its contingent counterpart as is seen in the second part of the proof above because the hypothetical assumption of the contingent property $A$ has to be exchanged for its necessitation for the proof structure to be preserved.

>>>>

\section{Anderson's Emendation in Natural Deduction}
\label{sec:anderson}

Anderson's solution \citep{and90} modifies the definitions of God-like being and essence, and eliminates half of Axiom 1 (replacing the equivalence by an implication).

@@

%Anderson modification of the axioms in A1 equivalence removed and definition of essence altered. 
%Anderson's implication instead of equivalence. 
\begin{axiom}[Emended]
\label{A1}
If the negation of a property is positive, the property is not positive:
$$
\all \varphi. [P(\neg \varphi) \imp \neg P(\varphi)]
$$
\end{axiom}

\begin{axiom}[As in Scott's]
\label{A2}
A property necessarily implied by a positive property is positive:
$$
\all \varphi. \all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)]
$$
\end{axiom}

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only small modifications: the equivalence of axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

\begin{definition}[Emended]
\label{D1}
A \emph{God-like} being necessarily possesses those and only those properties which are positive:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \biimp \nec \varphi(x)]
$$
\end{definition}

\begin{axiom}[As in Scott's]
\label{A3}
The property of being god-like is positive:
$$
P(G)
$$
\end{axiom}

\begin{corollary}
\label{C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof}
This proof is omitted, because it is identical to the proof of the same corollary shown in Sections \ref{sec:scott} and \ref{sec:newproof}.
\end{proof}

\begin{axiom}[As in Scott's]
\label{A4}
Positive properties are necessarily positive:
$$
\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
$$
\end{axiom}

\begin{definition}[Emended]
\label{D2}
An \emph{essence} of an individual is a property possessed by it and necessarily implying those and only those properties necessarily possessed by it:
$$
\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

\begin{definition}[As in Scott's]
\label{D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
$$
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex y.\varphi(y)]
$$
\end{definition}

\begin{axiom}[As in Scott's]
\label{A5}
Necessary existence is a positive property:
$$
P(E)
$$
\end{axiom}

Because of Anderson's emendations to Axiom 1 and Definition 1, the proof of Lemma 1 has to be modified as well. The proof shown below uses the modal axiom T to eliminate the extra box modality introduced by Anderson
in his emended Definition 1.

\begin{lemma}
\label{L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}
\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_6] \quad G(\gamma)$}
\end{prooftree}

@@

\begin{small}
\begin{prooftree}
\AXC{$\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\nec E(\gamma)$}
    \AXC{Axiom T} \dashedLine
    \UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

\setcounter{theorem}{2}
\begin{theorem}
\label{T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}
\begin{proof}
This proof is identical to the proof of Theorem 3 shown in Section \ref{sec:newproof}.
\end{proof}

We know from \citep{hajek} that Anderson's emendation also makes Axioms 4 and 5 derivable from the other axioms and definitions. Discussions about the redundancy of these two axioms started with an earlier (and wrong claim) claim by Magari \citep{magari} w.r.t. Scott's variant, and the redundancy was the topic of a long controversy \citep{controversy} between H\'ajek and Anderson. The proof presented above still uses Axioms 4 and 5, though, since the intention here was to show that Anderson's emendation requires only a minimal modification of the proofs presented in the previous sections.

ToDo: should we try to show here also an alternative proof of T3 using only Anderson's axioms A1, A2 and A3 and definition D1??  I don't know how hard this would be. Hájek1996 could be helpful.

\subsection{A Proof-Theoretical Explanation for the Immunity to Modal Collapse}

Anderson's change in the definition of essence adds an extra box modality, which transforms the standard proof (shown in Section \ref{sec:collapse}) of modal collapse (i.e. $\vdash A \imp \nec A$) into a proof of a tautology ($\vdash \nec A \imp \nec A$), as shown below.

@@

The exchange of the implication in Scott's definition of essence for an equivalence and the additional necessitation (box) modality before the property $\psi$ (which is instantiated by $\lambda x. A$) prevents the implication of a necessary fact from its contingent counterpart, as seen in the second part of the proof above, because the hypothetical assumption of the contingent property $A$ has to be exchanged for its necessitation $\nec A$ to preserve the proof structure.
